---
title: "Bayesian Time Series Analysis of US Unemployment Rates"
author: |
  | STAT 447C Final Project - Project Report
  |
  | Rohan Joseph (67089839)
date: "Due Date: April 19th 2024"
output: pdf_document
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(ggplot2)
library(readr)
library(tseries)
library(forecast)

```



# Key Details

- **Student(s):** Rohan Joseph (Student No.: 67089839)
- **Professor:** Alexandre Bouchard-Côté 
- **Course:** STAT 447C - Bayesian Statistics
- **GitHub Repository Link:** https://github.com/RohanUBC/STAT-447C-Final-Project





# Abstract

- Brief summary of objectives, methods, key results, focus on uncertainty quantification
and calibration.

[To be added]





# Introduction

- *Introduce the problem:* Forecasting US unemployment rates using Bayesian time series 
analysis.
- Provide context and the importance of understanding and predicting unemployment 
rates.
- The importance of forecasting and uncertainty quantification.
- *Context:* Briefly explain the significance of unemployment rates in economic 
planning and policy-making.

[To be added]



## Problem Formulation

The central focus of this project is to model and forecast unemployment rates in 
the United States (US), utilizing a real-world dataset that has documented the monthly
unemployment rate (from 01/1948 to 02/2024) in the US. A key element of the project 
is to accurately forecast these rates into the future while also quantifying the 
inherent uncertainty of such forecasts. The dataset that is being used is from the 
US Federal Reserve on monthly US unemployment rates.

*Federal Reserve Economic Data (FRED):* This dataset offer historical data on monthly 
US unemployment rates (from 1948 to 2024) and their associated economic indicators.

URL: https://fred.stlouisfed.org/series/UNRATE


We will adopt a Bayesian time series analysis approach, utilizing a Bayesian Structural
Time Series (BSTS) model to capture the dynamics of the US labor market based on the
provided data. This involves uncovering and modeling historical trends and seasonal 
patterns within the data, providing probabilistic forecasts of future unemployment rates.
We opt for BSTS due to its flexibility in modeling complex time series data with trend
and seasonality components. BSTS allows for the estimation of latent states, such 
as trend and seasonal effects, and provides a framework for forecasting future states
while quantifying uncertainty.


**Key Modelling/Methodological Challenge**

This project faces challenges in applying Bayesian time series analysis with BSTS
to economic data, particularly in modeling the seasonal patterns and long-term trends
inherent in unemployment rates. The primary hurdle lies in selecting and fine-tuning
parameters and models that can accurately reflect the dynamics of the US labor market.
This includes determining appropriate prior distributions, specifying the structural
components of the model, and validating the model's performance through diagnostics
and posterior predictive checks. Additionally, handling the irregularity in seasonal
variation presents a challenge, requiring careful consideration in the modeling approach
to ensure accurate representation of the data. 

In addition, we aim to also include a calibration of the uncertainty of the forecast,
demanding a sophisticated approach in how the model is constructed and validated. 
This is done to ensure that the forecasts are not only precise, but also accompanied 
by reliable uncertainty measures. Such measures are pivotal for economic planning 
and policy-making, providing a foundation for risk-aware decision-making. By accurately
quantifying forecast uncertainty, this approach aims to assess potential risks effectively, 
thereby facilitating more informed and prudent economic decision making.





## Literature Review

- Summarize key findings from existing literature on Bayesian forecasting and unemployment
rate analysis.
- Highlight gaps your project aims to fill or how it builds upon previous work.
- Discussion of uncertainty quantification in economic forecasting.
- Calibration of probabilistic forecasts in existing literature.

[To be added]





# Data Analysis

- *Model:* A Bayesian model is precisely described (e.g., using the ... ~ ... notation).
  - Description of BSTS for modeling unemployment rates.
- Implementation code in the appendix (e.g., using Stan)
  - R code for implementing the BSTS using Bayesian methods.
- *Motivation of prior choice:* If appropriate, several choices are compared or sensitivity
analysis is performed.
- *Critical evaluation of the posterior approximation:* An appropriate combination of
diagnostics, synthetic datasets and other validation strategies.
  - Techniques used for uncertainty quantification and calibration of forecasts.
  - Model diagnostics and validation strategies.
- *Methodological/Theoretical aspect:*
  - Assessment of the approach's robustness and creativity.
  - Discussion on the choice of BSTS and Bayesian calibration methods.
- *Calibration of uncertainty:* 
  - Explanation of calibration tests for predictive intervals.
  - Presentation of calibration test results.
  - Analysis of how well the model's uncertainty measures align with observed data.

[To be added]





# Discussion

- *Project Theme:* Assess the soundness and creativity of the approach within the context
of Bayesian forecasting and time series analysis. Discuss the model's ability to handle
the intricacies of unemployment data and its forecasting accuracy.
- *Summarizing results:* Summarize key findings and their implications for understanding 
and forecasting US unemployment rates.
  - Interpretation of the forecasting results.
- Implications of calibrated uncertainty for economic decision-making.
- Discuss the limitations of your study, such as data constraints, model assumptions,
or potential biases.
- Suggest areas for future research or methodological improvements.

[To be added]





# Conclusion

- Recap of the project's contributions to Bayesian forecasting and calibration.
- Reflect on the value of Bayesian methods in economic time series forecasting.
- Final thoughts on policy implications and the value of calibrated forecasting.

[To be added]





# References

[To be added]


\clearpage





# Appendix



## Appendix A - Explanatory Data Analysis


**Reading in the Dataset**

```{r}

dat <- read_csv("UNRATE.csv")
dat$DATE <- as.Date(dat$DATE)
head(dat)

```


**Historical US Unemployment Rates from 01/1948 to 02/2024**

```{r, fig.align = "center", fig.width = 7, fig.height = 4}

plot(dat$DATE, dat$UNRATE, 
     type = "o",
     pch = 20, 
     col = "red", 
     xlab = "Date", 
     ylab = "Unemployment Rate (in %)",
     main = "Figure 1: Historical US Unemployment Rate Over Time")
lines(dat$DATE, dat$UNRATE, col = "black")

```


**Average Yearly US Unemployment Rates from 01/1948 to 02/2024**

```{r, fig.align = "center", fig.width = 7, fig.height = 4}

avg_yearly_unrate <- dat |>
  mutate(YEAR = format(DATE, "%Y")) |>  
  group_by(YEAR) |>  
  summarise(Average_Unemployment_Rate = mean(UNRATE, na.rm = TRUE)) 

avg_yearly_unrate$YEAR <- as.numeric(as.character(avg_yearly_unrate$YEAR))

plot(avg_yearly_unrate$YEAR, avg_yearly_unrate$Average_Unemployment_Rate,
     type = "o", 
     pch = 20,
     col = "red", 
     xlab = "Year", 
     ylab = "Average Unemployment Rate (in %)",
     main = "Figure 2: Average Yearly Unemployment Rate in the US")
lines(avg_yearly_unrate$YEAR, avg_yearly_unrate$Average_Unemployment_Rate, col = "black")

```


\clearpage



## Appendix B - Time Series Analysis


**Decomposition of the Time Series**

```{r, fig.align = "center", fig.width = 7, fig.height = 6.5}

ts_unrate <- ts(dat$UNRATE, start = c(1948, 1), frequency = 12)
decompose_of_ts = stl(ts_unrate, s.window = "periodic")
plot(decompose_of_ts, 
     main = "Figure 3: Decomposition of the Time Series of US Unemployment Rates")

```

*Trend:* The trend plot in the decomposition suggests that there doesn't seem to
be any particular long-term trend in the US unemployment rates, however, there are
periods where the unemployment rate exhibits some trends, which seem to be non-linear
and varying with time.

*Seasonality:* The seasonal plot in the decomposition suggests that there is seasonal 
variation in the US unemployment rates, but there isn't a consistent interval in the
variation in the seasonality.


**Considering the Trend Component**

[To be added]


**Considering the Seasonality Component**

[To be added]


\clearpage



## Appendix [] - []

[To be added]



















