---
title: "STAT 447C Final Project - Project Proposal"
author: "Rohan Joseph (67089839)"
date: "Due Date: March 8th 2024"
output: pdf_document
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(dplyr)
library(ggplot2)
library(readr)
library(readxl)

```



# Key Details

- **Project Idea:** Bayesian Time Series Analysis of US Unemployment Rates 
- **Student(s):** Rohan Joseph (Student No.: 67089839)
- **Professor:** Alexandre Bouchard-Côté 
- **Course:** STAT 447C - Bayesian Statistics
- **Project Proposal Submittion Date:** March 15th 2024



# Project Theme and Overview

The aim of this project is to apply Bayesian forecasting techniques to model and 
predict unemployment rates in the United States (US), utlizing a real-world dataset
that has documented the monthly unemployment rate (from 01/1948 to 02/2024) in the 
US. Through the lens of Bayesian time series analysis, this project aims to reveal
seasonal patterns/fluctuations in the data, providing a detailed look at how the
unemployment rate has evolved over time, with a goal of providing probabilistic forecasts
of future trends. Furthermore, this project also aims to quantify the uncertainty
surrounding these forcasts, which is an important consideration to make when considering
the inherent variability over the years in the US labor market. 

This approach will enable a detailed assessment of the historical behavior of unemployment
rates, offering insights into the potential future state of the US labor market, 
solely based on its past trends. This analysis would hold significant value in understanding
US economic cycles, the planning of any governmental interventions, and for preparing
for future market conditions, while also showcasing how Bayesian methods can be applied
in economic time series forecasting.



# Data Sources

The two real-world candidate datasets that have been identified are:
 
**Federal Reserve Economic Data (FRED):** This dataset offer historical data on monthly 
US unemployment rates (from 1948 to 2024) and their associated economic indicators.

URL: https://fred.stlouisfed.org/series/UNRATE

```{r}

# The dataset was downloaded and saved locally and on the GitHub repository

FRED_dataset <- read_csv("UNRATE.csv")
head(FRED_dataset)

```

**US Bureau of Labor Statistics (BLS):** Just like the dataset from the US Federal 
Reserve, his dataset offer historical data on monthly US unemployment rates (from 1948
to 2024) and their associated economic indicators.

URL: https://data.bls.gov/pdq/SurveyOutputServlet

Please note that the original dataset downloaded from the above URL contains title
information, and so I have removed them when reading in the data in the console. 
However, I have saved the original dataset under the file name `BLS.xlsx`.

```{r}

# The dataset was downloaded and saved locally and on the GitHub repository

BLS_dataset <- read_excel("BLS_edited.xlsx")
head(BLS_dataset)

```

These two datasets can be used well to tackle the selected project theme.



# GitHub Repository Link

The following link takes you to the public GitHub repository where the project files
are located.

URL: https://github.com/RohanUBC/STAT-447C-Final-Project

The repository includes the following files:

- Project Proposal
- Project Report (scaffold)
- The Federal Resereve Economic Data (FRED) Dataset
- The US Burean of Labour Statistics (BLS) Dataset
- A README File 




